---
author: Jack Williams and Quinn Asena
---

# Age-depth modelling {#sec-age-depth}

## Background

A foundational difference between geology and ecology is that, for geologists, time is an unknown variable that must be estimated with uncertainty. In contrast, most ecologists can assume that the temporal coordinates of their observations are known precisely, with zero uncertainty. Fortunately, geochronologists have a wide variety of natural clocks, thanks to the constant decay rates of radioactive isotopes. Each isotope has a unique decay rate, and so each is appropriate for particular timescales.

For the last 50,000 years, radiocarbon (^14^C), with its half-life of 5,730 years, is by far the most common form of radiometric dating. (Beyond 10 half-lives, so much of a radioactive substance has decayed away that it becomes immeasurable.) Radiocarbon is the mainstay of Quaternary dating and archaeology.

In Quaternary paleoecology, radiocarbon dating is expensive , a single sample typically costs \$300 to \$500, so usually a given lake-sediment record will have only a scattering (ca. 5 to 30) of radiocarbon dates and other age controls. Other kinds of age controls include volcanic ash layers (tephras), ^210^Pb (half-life: 22.3 yrs), optically stimulated luminescence (OSL) dates, historic events such as the rise in *Ambrosia* pollen abundances associated with EuroAmerican land clearance, etc. An age-depth model must be constructed to estimate the age of sediments not directly associated with an age control. In multi-site data syntheses, the number of age controls, their precision, and their likely accuracy are all fundamental indicators of data quality [e.g. @blois2011; @mottl2021].

To estimate ages for depths lacking radiocarbon date, an age-depth model is required. Age-depth models are fitted to the available age controls (age estimates with uncertainty for individual depths) and provide estimates of age as a function of depth, for all depths and ages within the temporal bounds of the model.

## Getting started

Here we will gain practice in working with age-depth models of various kind, and assessing their age estimates and uncertainty. We’ll begin with a bit of practice in calibrating radiocarbon years to calendar years and comparing the resulting age estimates from different calibration curves.


::: {.callout-note}
# Packages required for this section

We will be using `Bchron` for calibration and Bayesian age-depth modelling. Notably `rbacon` is another commonly used package, see @trachsel2017 for a discussion on age-depth models. We will also be fitting some interpolation and linear models using base R. Remember you must have packages both installed _and_ loaded, often loading a package is done via the `library(packagename)` function. The following code using the `pacman` package takes care of installation and loading in one go.

```{r ageDepthPackages}
#| output: false
# Load up the package
if (!require("pacman")) install.packages("pacman", repos="http://cran.r-project.org")
pacman::p_load(Bchron, splines, ggplot2) # splines comes with base R
```
:::


### Calibration of radiocarbon dates

A complication in radiocarbon dating is that the initial calculation of a radiocarbon age assumes, by convention, that the amount of radiocarbon in the atmosphere is constant over time. See @bronkramsey2008 for a good overview of ^14^C dating. This assumption is untrue, so all radiocarbon age estimates must be post-hoc calibrated using a calibration curve that is based on compiling radiocarbon dates of materials that have precise independent age estimates (e.g. tree rings, corals). The IntCal series (IntCal04, IntCal09, IntCal13, IntCal20) is the community standard for calibrating radiocarbon dates to age estimates in calendar years [e.g., @reimer2020]. The conversion from radiocarbon to calendar years usually further increases the uncertainty of age estimates.

Yet another complication in radiocarbon dating is that different calibration curves need to be used for the Northern vs. Southern Hemisphere and for the atmosphere vs. oceans, due to different residence times of ^14^C in these different reservoirs. For example, atmospheric radiocarbon that diffuses into the surface oceans usually will reside for centuries before phytoplankton biologically fix it through photosynthesis, which will lead  marine ^14^C to be depleted (and ‘too old’) relative to atmospheric ^14^C. Use the wrong calibration curve and your age estimate will be inaccurate!


Before calibrating several dates at once, it is worth looking at just one, together with the
curve it was calibrated against. The shaded distribution on the left is the calibrated calendar
age; the diagonal band is the calibration curve itself.

```{r fig-single-date}
#| fig-cap: "One radiocarbon date of 11,553 +/- 50 14C BP, calibrated against IntCal20, shown with the calibration curve it was matched to."
BchronCalibrate(ages = 11553,
                ageSds = 50,
                calCurves = "intcal20") %>%
  plot(includeCal = TRUE)
```

Notice that the calibrated distribution is not symmetrical, and not necessarily a single peak.
Its shape is inherited from the wiggles in the curve, which is why a calibrated age can rarely
be summarised honestly by one number.

#### Calibrating radiocarbon dates in R

Here we’ll experiment with calibrating radiocarbon dates, using various calibration curves. Radiocarbon dated samples come back from the lab with a radiocarbon age and standard deviation, among other information. These two bits of information are used to calibrate the radiocarbon dates to estimated ages. R packages may have useful vignettes (package tutorials) and built-in datasets that provide handy test templates. The following code is modified from the `Bchron` [vignette](https://cran.r-project.org/web/packages/Bchron/vignettes/Bchron.html).

::: {.panel-tabset}

## Code

```{r bchronCalibrate}
#| output: false
ages <- BchronCalibrate(ages = c(3445, 11553, 7456),
                        ageSds = c(50, 230, 110),
                        calCurves = c('intcal20', 'intcal20', 'intcal20'))
```

## Summary

```{r bchronCalibrateSummary}
summary(ages)
```

## Plots

```{r bchronCalibratePlot}
plot(ages)
```

## Calibration curve plots

```{r fig-calibration-curves}
#| fig-cap: "Three calibrated dates shown against the IntCal20 calibration curve."
plot(ages, includeCal = TRUE, fillCol = 'red')
```

:::

The output summary indicates the range of the highest density regions (i.e., the most likely 'real' age range of the sample). The `plot` function in `Bchron` outputs a `ggplot` object of the high density regions of the most likely ages.

The object you assigned the result of `BchronCalibrate()` to holds all of the calibration
information, and it can be reached with `$`, for example `ages$Date1$ageGrid`.

::: {#exr-cal-curves}
Using the example code above, change the `calCurves` input to:

1. the Southern Hemisphere calibration curve,
2. the marine sediments curve.

How much do the estimated ages move?
:::

::: {#exr-cal-different-ages}
Calibrate a different set of three ages by modifying the inputs to `ages` and `ageSds`.
:::

::: {#exr-cal-median}
Using one of your calibrated sets of three dates, calculate the median age from the `ageGrid`,
and the difference between each calibrated age and the original uncalibrated age.
:::

::: {.callout-note}
# Median and mean ages

Commonly, only the median age is reported; however, it is good practice to report and consider
the range of possible ages, especially when exploring synchroneity of events across space.
Notice how `Bchron` focuses on delivering ranges and densities.

:::

### Types of age-depth model

Different kinds of age-depth models exist, each with their own underlying assumptions and behaviour. In the list below, #1-4 are classical or traditional forms of age-depth models, but Bayesian models are now the norm. The packages `rbacon` (usually referred to as 'bacon') and `Bchron` are the current standards for Bayesian age-depth modelling. Before going to Bayesian models, we’ll begin with the classics.

1. **Linear interpolation**, a.k.a. ‘connect the dots,’ in which straight lines are drawn between each depth-adjacent pair of age controls.

2. **Linear regression** ($y = b_0 + b_1 x$, where $y$ is age, $x$ is depth, and $b_0$ and $b_1$ are constants), in which a single straight line is fitted through the entire set of age controls. In ordinary least squares (OLS), the chosen line is the one that minimises the sum of the squared vertical distances between the points and the line. Standard OLS assumes that all observations are normally distributed, which is a poor assumption for calibrated radiocarbon dates.

3. **Polynomials**, also fitted to the entire set of age controls ($y = b_0 + b_1 x + b_2 x^2 + b_3 x^3 + \dots + b_n x^n$), are an extension of linear regression, with additional terms for $x^2$, $x^3$, etc. Some arbitrary maximum $n$ is chosen, usually in the range of 3 to 5. These are called 'third-order polynomials,' 'fifth-order polynomials,' etc.

4. **Splines**, which are a special kind of polynomial function that are locally fitted to subsets of the age controls, and then smoothly interpolated between points. (Several different formulas can be used to generate splines; common forms include cubic, smoothing and monotonic splines. LOWESS is often used for the same purpose but is a local regression smoother rather than a spline.)


5. **Bayesian age models** (e.g. `bacon`, `bchron`, `oxcal`, etc.). Bayesian models differ in detail, but all combine a statistical representation of prior knowledge with the new data (i.e. the age controls at a site) to build an age-depth model with posterior estimates of age for any given depth. Bayesian models are now widely used because

    1. they allow the incorporation of prior knowledge (e.g., from coring many lakes, we now have decent estimates of typical sediment accumulation rates, @goring2012)
    2. they can handle highly irregular probability distribution functions such as those for radiocarbon dates after calibration; and as a result
    3.  they generally do a better job of describing uncertainty than traditional age-depth models


#### Classical age-depth models

Classical models are now out-dated methods [@blaauw2018], but it is useful to understand how they work as literature before the relatively recent development of Bayesian methods has relied on them. Let's explore some classical methods of age-depth modelling using one of the datasets included with the `Bchron` package. The data are from a core at Sluggan Moss in County Antrim, Northern Ireland. The published record calls the site Sluggan Bog [@smith1991]; both names refer to the same place, and `Bchron` uses both in its own documentation. The dataset is called via:

```{r slugganData}
data(Sluggan) # Call the data from Bchron
head(Sluggan) # Check out the data
```

Notice that some depths appear more than once. Sluggan Moss was dated in replicate at several
levels, with three separate radiocarbon determinations at 272.5 cm and at 367.5 cm, for
instance. Replicate dating is good practice, because it shows how reproducible a determination
is, but it does mean the classical methods below have to decide what to do when one depth
carries several ages.

::: {.panel-tabset}

## Linear interpolation

Linear interpolation predicts ages by simply, drawing a line between successive dated samples. This method assumes that there is a constant age-depth relationship between samples. An assumption that is unlikely to be true, especially for cores with fewer dated samples than Sluggan Moss.

```{r fig-linear-interp}
#| fig-cap: "Linear interpolation, drawing straight lines between successive dated levels."
# Interpolate an age for every depth. ties = mean averages the replicate dates at
# those depths that were dated more than once.
interp_ages <- approx(x = Sluggan$position, y = Sluggan$ages, ties = mean)

# Age on the horizontal axis, depth on the vertical and increasing downwards
plot(x = interp_ages$y, y = interp_ages$x, type = "l",
     xlab = "Age (14C years BP)", ylab = "Depth (cm)",
     ylim = rev(range(Sluggan$position)))
points(x = Sluggan$ages, y = Sluggan$position, cex = 1.5, col = "red")
```

## Linear regression

Linear regression provides a line of best fit through the dated samples. This method assumes a constant age-depth relationship across all samples, also unlikely to be true depending on processes affecting the core during its formation.

```{r fig-linear-reg}
#| fig-cap: "Linear regression, a single straight line fitted through every age control."
mod_ages <- lm(ages ~ position, data = Sluggan) # age as a function of depth

# A grid of depths to predict ages at, used by this model and the two below
depth_grid <- seq(from = min(Sluggan$position), to = max(Sluggan$position),
                  length.out = 250)

plot(x = Sluggan$ages, y = Sluggan$position, cex = 1.5, col = "red",
     xlab = "Age (14C years BP)", ylab = "Depth (cm)",
     ylim = rev(range(Sluggan$position)))
lines(x = predict(mod_ages, data.frame(position = depth_grid)), y = depth_grid)
```

## Polynomial regression

Polynomial regression allows a curve to be fit through the data. The amount the curve 'wiggles' depends on the order of the polynomial fit to the data. Polynomial regression has the risk of being over-fit.

```{r fig-poly-reg}
#| fig-cap: "A third-order polynomial fitted to the age controls."
poly_ages <- lm(ages ~ poly(position, 3), data = Sluggan)

plot(x = Sluggan$ages, y = Sluggan$position, cex = 1.5, col = "red",
     xlab = "Age (14C years BP)", ylab = "Depth (cm)",
     ylim = rev(range(Sluggan$position)))
lines(x = predict(poly_ages, data.frame(position = depth_grid)), y = depth_grid)
```

## Cubic splines

Splines are a class of functions including, for example, smoothing splines and cubic splines.
Cubic splines are piece-wise polynomials fitted locally between 'knots'. That is, the data are
split into bins, and a separate polynomial is fitted within each bin, with the pieces joined
smoothly at the knots. By default the `bs()` function uses a third-degree polynomial, so
without any knots the fit looks the same as the third-order polynomial above. Note that the
knots are positions along the predictor, so here they are depths in centimetres.

```{r fig-cubic-spline}
#| fig-cap: "A cubic spline with knots at 100, 250 and 400 cm depth."
cubic_ages <- lm(ages ~ splines::bs(position, knots = c(100, 250, 400)),
                 data = Sluggan)

plot(x = Sluggan$ages, y = Sluggan$position, cex = 1.5, col = "red",
     xlab = "Age (14C years BP)", ylab = "Depth (cm)",
     ylim = rev(range(Sluggan$position)))
lines(x = predict(cubic_ages, data.frame(position = depth_grid)), y = depth_grid)
```

:::

Because classical age-depth modelling is rarely used now we are not going to delve further into the statistical details of the best way of fitting each model (e.g., the number of knots to use for fitting a cubic spline).

One of the issues with classical age-depth modelling is that uncertainty _decreases_ with fewer datapoints.

::: {#exr-glendalough-classical}
`Bchron` has a second built-in dataset named Glendalough, which has far fewer radiocarbon
dates than Sluggan Moss. Using the Glendalough data:

1. fit a linear interpolation, a linear regression, three polynomials of different order, and
   a cubic spline,
2. plot the results.

Did you run into any errors or warnings? If so, why?
:::


#### Bayesian age-depth models

Now let's see what the latest methods show for the same dataset.

::: {.panel-tabset}

## Code

Note that all the values provided to the arguments are contained in the Sluggan dataframe. When creating chronologies from your own (or accessed data), you may need to rename them to match your data.

```{r bayesianModel}
#| output: false

set.seed(1984) # the model is stochastic, so fix the seed for reproducibility

SlugganOut <- with(Sluggan,
                   Bchronology(ages = ages,
                               ageSds = ageSds,
                               calCurves = calCurves,
                               positions = position,
                               positionThicknesses = thickness,
                               ids = id,
                               predictPositions = seq(min(Sluggan$position),
                                                      max(Sluggan$position),
                                                      by = 10)))

```

## Summary

The summary shows for each position (depth) the median and quartiles of the predicted ages for that position.

```{r bayesianSummary}
summary(SlugganOut)

```

## Outliers

`Bchron` estimates, for each date, the probability that it is an outlier given all of the
others. A high value flags a date that disagrees with its neighbours.

```{r bayesianOutliers}
summary(SlugganOut, type = "outliers")
```

## Plot

```{r fig-sluggan-chronology}
#| fig-cap: "The Bchron age-depth model for Sluggan Moss. The shaded envelope is the posterior age range at each depth, and it narrows close to a dated level and widens between them."
plot(SlugganOut)
```

:::

A more complete version of the modelling process using Bchron can be found in the [vignette](https://cran.r-project.org/web/packages/Bchron/vignettes/Bchron.html).

::: {#exr-glendalough-bayesian}
Run the `Bchron` age-depth model on the Glendalough dataset, then consider:

1. how does the uncertainty shown in the plot compare between Glendalough, which has few
   radiocarbon dates, and Sluggan Moss, which is data-rich?
2. run `summary()` on the Glendalough model. Does the output reflect what the plot shows? Is
   the uncertainty smaller close to a dated sample? Why does interpreting a result in the
   context of its uncertainty matter?
:::

::: {.callout-tip}
# Resources

Age-depth modelling is complicated, there are many pitfalls, assumptions, and uncertainties that are often ignored. Recent developments have begun to focus on quantifying uncertainties to understand the reliability of inferences made from the data. Key papers for understanding age depth modelling include:

- @blaauw2018

- @parnell2008

- @trachsel2017

:::

### Accumulation rate

Having a chronology is not the end of it. Because the model gives an age at every depth, it
also tells us how fast sediment accumulated, which is simply the rate of change of age with
depth. `Bchron` will return that directly, as a distribution rather than a single line:

```{r accRate}
#| output: false
acc_rate <- summary(SlugganOut,
                    type = "acc_rate",
                    probs = c(0.25, 0.5, 0.75))
```

```{r fig-accumulation-rate}
#| fig-cap: "Sediment accumulation rate through the Sluggan Moss core. The line is the median and the ribbon the interquartile range."
#| warning: false
ggplot(acc_rate, aes(x = age_grid, y = `50%`, ymax = `75%`, ymin = `25%`)) +
  geom_line() +
  geom_ribbon(alpha = 0.2) +
  scale_x_reverse() +
  labs(y = "Accumulation rate (cm per year)",
       x = "Age (cal years BP)") +
  theme_bw()
```

Two things are worth drawing out of @fig-accumulation-rate. Accumulation rate is a derivative
of the age-depth model, so it inherits all of that model's uncertainty and amplifies it; the
ribbon here is an interquartile range, not a measurement error. And the rate is not constant,
which is precisely the assumption that linear interpolation and linear regression were making
earlier in this chapter. Sediment does not accumulate at a steady pace, and a chronology that
pretends otherwise will misdate everything between its control points.

## Conclusion

We have calibrated radiocarbon dates and seen how much uncertainty that conversion adds, fitted
the four classical age-depth models and seen where each of them misleads, and built a Bayesian
chronology that carries its uncertainty through to every predicted age. The chronology is the
foundation everything else in this book rests on: every rate, every ordination and every
comparison between sites inherits the uncertainty established here.

Along the way there were five exercises. They are worth returning to if you skipped any, because
the habit they build, of asking how well an age is really known, is the one that matters most:

- @exr-cal-curves, choosing the right calibration curve
- @exr-cal-different-ages, calibrating your own dates
- @exr-cal-median, and why a median is a poor summary of a calibrated date
- @exr-glendalough-classical, the classical models on a sparse record
- @exr-glendalough-bayesian, and how uncertainty grows between dated levels

