---
author: Quinn Asena
---

# Visualising data {#sec-visualisation}

## Background {#sec-visualisationbackground}

One of the most common methods of plotting palaeoecological data (such as pollen, but often including diatoms, isotopes, and other palaeoecological data) is to create stratigraphic plots. Palaeodata are regularly visualised with age (usually the output of an age-depth model) on the y-axis, with the youngest age at the top, to mirror a core-type sample. Each taxon is plotted in sequence horizontally, in order of abundance. Taxonomic data, such as pollen or diatoms, are also usually converted to relative abundances before plotting. While the relative abundances do help visualise the proportion of each taxon through time (and absolute abundances are unavailable), this also means that the plots can be misleading. For example, if all taxa were to decline equally, then the figure would not change at all! Or the rise of a taxon could be due to the decline of others. Of course, this is not the only method of visualising palaeoecological data, and other layouts can be equally good; however, we will demonstrate the most common convention in this chapter. We will do so using the `ggplot2` package and the `rioja` package.


## Getting started

::: {.callout-note}
# Packages required for this section

```{r visPackages}
#| output: false
if (!require("pacman")) install.packages("pacman", repos = "https://cran.r-project.org")
pacman::p_load(dplyr, tidyr, tibble, ggplot2, scales, stringr, mgcv, rioja, neotoma2, marquee)
```
:::

::: {.callout-tip}
# Respawn code

This is the same Devil's Lake download used in @sec-age-depth and @sec-dissimilarity, filtered
to `datasettype == "pollen"`, harmonised the same way, and converted to proportions.

```{r visDevilsDownload}
#| output: false
if (!file.exists("./data/devils_dl.rds")) {
  devils_dl <- get_sites(siteid = 666) %>%
    get_datasets(all_data = TRUE) %>%
    neotoma2::filter(datasettype == "pollen") %>%
    get_downloads()
  saveRDS(devils_dl, "./data/devils_dl.rds")
} else {
  devils_dl <- readRDS("./data/devils_dl.rds")
}

```
:::


Most stratigraphic plots you will see have taxa converted to relative abundances. The following code converts observations from counts (NISP) to proportions. In this section we will not go into detail on the `samples()` function or the conversion to proportions as this is covered in @sec-neotoma2 and @sec-dissimilarity.

```{r ConvertToProp}
devils_samples <- samples(devils_dl) %>%
  dplyr::filter(ecologicalgroup %in% c("UPHE", "TRSH"),
                elementtype == "pollen",
                units == "NISP")

devils_harm_prop <- devils_samples %>%
  mutate(variablename = replace(variablename,
                                stringr::str_detect(variablename, "Pinus.*"),
                                "Pinus")) %>%
  group_by(siteid, sitename, sampleid, variablename, units,
           age, agetype, depth, datasetid, long, lat) %>%
  summarise(value = sum(value), .groups = "drop") %>%
  group_by(age) %>%
  mutate(pollencount = sum(value, na.rm = TRUE),
         prop        = value / pollencount) %>%
  ungroup() %>%
  arrange(desc(age))
```



## Stratigraphic diagrams

In most publication-ready figures, you would plot as many taxa and functional groups as possible. For this demonstration we will plot the 6 most abundant species so that they fit more easily on the rendered page. We will do this simply by summing the taxa through time and picking the top 6:

```{r TopTaxa}
devils_top_taxa <- devils_harm_prop %>%
  group_by(siteid, sitename, variablename) %>%
  summarise(total = sum(value), .groups = "drop") %>%
  arrange(desc(total))
```

Of course, this is an over-simplification since ecologically important does not necessarily also mean most abundant! It serves the purpose of having something to plot.

```{r visFocalTaxa}
focal <- devils_top_taxa$variablename[1:6]
# Filter to include only top 6 taxa
devils_focal <- devils_harm_prop %>%
  dplyr::filter(variablename %in% focal)

```

### Building a stratigraphic plot with `ggplot2`

Let's start with the simplest thing that works: proportion against age, one panel per taxon.
`coord_flip()` puts age on the vertical axis and `scale_x_reverse()` makes age run downwards, so
the oldest sample sits at the bottom. `facet_wrap()` splits each taxon (`variablename`) into one panel per taxon. We will incrementally build up from a basic plot to a fleshed-out example with correct details. If you are unfamiliar with ggplot, try commenting out or removing lines systematically, rerunning the code each time to see the difference.

```{r fig-strat-first}
#| fig-cap: "A first pollen diagram for Devil's Lake. Incomplete, but something to build on."
#| fig-height: 5
ggplot(devils_focal, aes(x = age, y = prop)) +
  geom_area(fill = "grey25") +
  scale_x_reverse() +
  scale_y_continuous() +
  labs(x = "Age (cal years BP)", y = "Pollen proportions") +
  coord_flip() + # confusingly age is the x label because of coord_flip()
  facet_wrap(~variablename, nrow = 1) + # makes each taxa its own panel
  theme_minimal()
```

Now let's start improving on this base plot @fig-strat-first! Using a filled area is common; however, the fill implies a continuous record and the core was sampled at discrete intervals. Everything between two samples is drawn by the plotting function, not measured. Adding a segment
at each sampled age shows where the observations actually are. The density of observations through time is very important, and it is a common mistake not to include them in stratigraphic plots! Currently, the panels are also in alphabetical order by default. Ordering by abundance is tidier and conveys relative abundance along the horizontal alignment of panels. To do this `ggplot2` likes to have the taxa as factor data, so let's convert the taxa names to factor data:

```{r visFactorOrder}
# convert taxa to factor, see <fct> now appears in the
# variablename column if you print the tibble?
devils_focal <- devils_focal %>%
  mutate(variablename = factor(variablename, levels = focal))

```

Now rebuild the plot with `geom_segment()` to add the sampling locations, and rebuild with the `variablename` as a factor:

```{r fig-strat-second}
#| fig-cap: "The same diagram with a segment marking every sampled level, and the panels ordered by total abundance rather than alphabetically."
#| fig-height: 5
ggplot(devils_focal, aes(x = age, y = prop)) +
  geom_area(fill = "grey25") +
  geom_segment(aes(x = age, xend = age, y = 0, yend = prop),
               colour = "grey40", linewidth = 0.3) +
  scale_x_reverse() +
  scale_y_continuous() +
  labs(x = "Age (cal years BP)", y = "Pollen proportions") +
  coord_flip() + # confusingly age is the x label because of coord_flip()
  facet_wrap(~variablename, nrow = 1) + # makes each taxa its own panel
  theme_minimal()
```

Now we're getting somewhere!


#### Formatting taxon names

By convention genus and species names are italic, the genus is capitalised and the species is not (e.g., *Quercus alba*). The `marquee` package renders markdown inside `ggplot2` theme elements, so a label is written the way it would be written using [markdown syntax](https://www.markdownguide.org/basic-syntax/#emphasis). Markdown is a featherweight plain text format that is super easy to use and commonplace in programming. There are many, many ways of achieving the same aim in `ggplot2`; I find markdown elements the simplest.

Let's create some labels for `ggplot2`. Since there are only 6 taxa, we can easily enough create matching pairs manually:

```{r visTaxonNames}
taxon_names <- c(
  Quercus  = "*Quercus* spp.",
  Pinus    = "*Pinus* spp.",
  Picea    = "*Picea* spp.",
  Ulmus    = "*Ulmus* spp.",
  Fraxinus = "*Fraxinus* spp.",
  Betula   = "*Betula* spp."
)

# raise an error if anything changes and taxon names do not match.
stopifnot(all(focal %in% names(taxon_names)))
```

Alternatively, we can take a simple programmatic approach:

```{r visTaxonNamesprogrammatic}
markdown_names <- paste0("*", focal, "* spp.")
taxon_names_prog <- setNames(markdown_names, focal)
stopifnot(all(focal %in% names(taxon_names_prog)))
```

For more complex lists of species names, you can take a rule-based approach with lots of regular expressions (regex) to find and match patterns like `undiff.`, `-type`, and `subg`. Pattern matching text got much easier with the help of artificial intelligence, so generating labels is not a hard task now (it used to be challenging to safely/correctly match all the tricky taxon names as well as isotopes with the correct notation!). 

Remember that the taxa are split into facets? Thus, it is the facet strip labels that need to be modified. `as_labeller()` puts the formatted strings into the facet strips, and `element_marquee()` tells `ggplot2` to render the markdown rather than print it. Both are needed: the labeller alone would
show the asterisks. Now the same plot with the segments, the ordering and the names:

```{r fig-strat-devils}
#| fig-cap: "Devil's Lake pollen diagram. Each horizontal segment is a sampled level, so the width of the shaded area between segments is interpolation rather than data. Panels are ordered by the total abundance of the taxa."
#| fig-height: 5
ggplot(devils_focal, aes(x = age, y = prop)) +
  geom_area(fill = "grey25") +
  geom_segment(aes(x = age, xend = age, y = 0, yend = prop),
               colour = "grey40", linewidth = 0.3) +
  scale_x_reverse(name = "Age (cal years BP)", labels = label_comma()) +
  scale_y_continuous(name = "Proportion of the pollen sum",
                     labels = label_percent(), n.breaks = 3) +
  coord_flip() +
  facet_wrap(~variablename, nrow = 1,
             labeller = as_labeller(taxon_names)) +
  theme_minimal() +
  theme(strip.text        = element_marquee(),
        panel.spacing.x   = unit(0.9, "lines"),
        panel.grid.minor  = element_blank(),
        axis.text.x       = element_text(size = 7))
```


Notice that we also included some optional cosmetic changes like percent on the x-axis using `label_percent()` and `label_comma()` to the ages on the y-axis. If you need to place multiple ggplot objects in the same figure, the [`patchwork`](https://patchwork.data-imaginist.com) package works very well with `ggplot2`.

::: {#exr-freeaxis}
1. `facet_wrap` includes optional arguments `"fixed"`, `"free"`, `"free_x"`, and `"free_y"`. Experiment with the alternative arguments to `facet_wrap` and discuss the consequences of how proportions are visualised.
:::

::: {.callout-note}
# Assumptions of stratigraphic plots of taxa :mage:

- When plotting relative abundances, that the taxa are _not_ independent of each other. As noted in @sec-visualisationbackground, a perceived increase in one taxon may be driven by the decrease in other species.

- Chronological uncertainties are not visually accounted for. @sec-age-depth covers how important the age-depth uncertainties can be and that only considering the median density can lead to incorrect conclusions. Common errors in palaeoecology are to: (i) infer the timing of ecological events from stratigraphic diagrams that do not account for chronological uncertainty; and (ii) plot data from multiple cores against a single chronology to infer the synchroneity of events across space without due consideration of the compounded chronological uncertainties. Although these considerations are important to visualisation and interpretation, we will not attempt to tackle such big issues in this chapter. If you are interested, here are a few papers:
  - @blois2011
  - @parnell2008
  - @blaauw2007
  - @blaauw2020

:::

::: {#exr-downsample}

Take Devil's Lake, or another site of your choosing.

1. Rebuild the chronology with `Bchron`, following @sec-age-depth, three times: once with the complete chronological control data, which is your reference point, and twice more with the control data artificially downsampled by removing rows. Keep the first and last control in each downsampled version.

2. Artificially downsample the pollen data twice, again by removing rows. The complete dataset is your reference point here too.

3. You now have three chronologies and three pollen datasets. Plot combinations of the two to see the effect of reduced temporal resolution in the chronological controls and in the pollen separately.
:::


### Stratigraphic diagram with `rioja`

The `rioja` package is built for visualising palaeoecological data and has some handy features like including CONISS [@grimm1987] constrained cluster analysis that splits the record into zones (we will talk more about calculating dissimilarity/similarity in @sec-dissimilarity). While `rioja` has useful features, it is an older package. In this section we will demonstrate how it works, but the chapter will primarily focus on using `ggplot2` which is the most frequently used graphics package in R and is actively maintained.

Two things are worth knowing before reading the code. Firstly `rioja` draws in base graphics rather than `ggplot2`, so the syntax is not transferable. For example italics come from a plotmath expression rather than the markdown `marquee` understands, and the two are not interchangeable. Secondly, `rioja` places the plot using figure coordinates running from 0 to 1, rather than the margins base graphics normally uses, so `xLeft`, `xRight` and `yTop` are what you adjust when something is cut off at an edge. Unlike `ggplot2` which usually likes to work with long data, `rioja` takes wide format. So, let's pivot our data to wide format:


```{r visWideTable}
#| output: false
devils_wide <- devils_harm_prop %>%
  dplyr::select(age, variablename, prop) %>%
  dplyr::filter(variablename %in% focal) %>%
  tidyr::pivot_wider(id_cols = age,
                     names_from  = variablename,
                     values_from = prop,
                     values_fill = 0)

head(devils_wide)
```

```{r fig-strat-rioja}
#| fig-cap: "The same six taxa drawn by `rioja::strat.plot()`, with a CONISS dendrogram on the right and four zones marked in red. `wa.order` sorts the taxa by weighted average age, so the panel order differs from the previous figure."
#| fig-height: 6
#| warning: false

# calculate clusters using CONISS. This is over-simplified to 6 species for this demonstration!
devils_clust <- rioja::chclust(dist(sqrt(devils_wide[, -1])), method = "coniss")

devils_matrix <- as.data.frame(devils_wide[, -1] * 100)

# rioja draws in base graphics, so the labels are plotmath expressions rather than
# the markdown marquee used above. Supply them in the ORIGINAL column order:
# wa.order reorders the panels and carries the labels along with them.
taxon_expr <- parse(text = paste0("italic('", names(devils_matrix), "')~'spp.'"))

strat <- rioja::strat.plot(devils_matrix,
                           yvar     = devils_wide$age,
                           ylabel   = "Age (cal years BP)",
                           x.names  = taxon_expr,
                           y.rev    = TRUE,
                           clust    = devils_clust,
                           wa.order = "topleft",
                           scale.percent = TRUE,
                           # everything below is layout. rioja positions the plot in
                           # figure coordinates, 0 to 1, rather than using margins.
                           xLeft       = 0.12,  # left edge, room for the y label (default 0.07)
                           xRight      = 0.94,  # right edge, room for the dendrogram (default 1)
                           yTop        = 0.75,  # top edge, room for the rotated names (default 0.8)
                           xSpace      = 0.025, # gap between panels (default 0.01)
                           ylabPos     = 3.2,   # y label distance from the axis (default 2)
                           clust.width = 0.12,  # width of the dendrogram (default 0.1)
                           cex.ylabel  = 0.9,
                           cex.xlabel  = 0.85,
                           cex.yaxis   = 0.7,
                           cex.axis    = 0.65)

# identify zones from CONISS clustering
rioja::addClustZone(strat, devils_clust, 4, col = "red")
```

It can be a bit fiddly to correct the layout in `rioja`, but this is also true of `ggplot2` when figures get complicated. Drawing plots entirely from data can be time consuming but has the benefit of reproducibility.

::: {.callout-tip}
# Good practice tip

You will want to save your figures to disk after they have been created. The graphics viewer in your IDE (e.g., RStudio or VS Code) does not display figures on-screen the same as they will appear when saved to disk. If you screenshot or copy from the graphics console, you will not be able to create the same figure dimensions twice and will have lots of problems downstream! Here are some tips for creating good, publication-ready, graphics:

- Define overall figure dimensions: this is easily done with [`ggsave()`](https://ggplot2.tidyverse.org/reference/ggsave.html) by defining the `width` and `height` arguments. Defining the correct figure size for a page or a slide will mean that you don't need to resize it (shrinking or blowing up fonts and other elements) when you place them in a document or presentation.
- Define all font sizes. These are easily fixed in `ggplot2` [`theme` section](https://ggplot2.tidyverse.org/reference/theme.html). That way, on the printed page you can make sure labels are visible and consistent at a chosen font size (e.g., axis labels will be font size 10 across all your figures). You can make many other cosmetic adjustments in the `theme` too.
- Save plot as a vector graphic format like `svg` or `pdf`, not as a raster format like `jpeg` or `png` unless absolutely necessary. Vector formats are much sharper and clearer. And if you need to adjust the figure in another software like Inkscape, each element will be a selectable/editable layer (a raster graphic will appear as like a photograph in the editor).
- It is worth using a colour palette designed for contrast and colourblind friendliness. [`RColorBrewer`](https://r-graph-gallery.com/38-rcolorbrewers-palettes.html) and [`viridis`](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html) can be aesthetic as well as functional.

:::


## Taxa across space and time

The above sections cover creating a stratigraphic diagram for a single site. Don't forget that the data downloaded from `neotoma2` contain many, many (many!) variables that can be used for data exploration. Let's have a look at a couple of other exploratory figures we could make from the multi-site object covering all of the state of Michigan downloaded in @sec-neotoma2. One site gives a history. Many sites give a geography as well. For demonstration purposes, we will load the data, filter for only records containing pollen data, and do a quick-and-dirty harmonisation for pine species:

```{r visMichSamples}
mich_dl <- readRDS("./data/mich_dl.rds")
```


### Taxa distribution and rarity

The `taxa()` function summarises the taxa across sites in one table: for every taxon, how many sites and how many samples it turns up in. Let's explore how we can use the taxa table to produce exploratory plots of the multisite data. First, we need to apply the same filters to the taxa table as to the samples table:

```{r visMichTaxa}
mich_taxa <- neotoma2::taxa(mich_dl) %>%
  dplyr::filter(ecologicalgroup %in% c("UPHE", "TRSH"), elementtype == "pollen")

cat("Total taxa: ", nrow(mich_taxa), "\n",
    "Number of taxa that occur only in one site: ", sum(mich_taxa$sites == 1), "\n",
    "Number of taxa that occur in every site: ", sum(mich_taxa$sites == max(mich_taxa$sites)), "\n")

```

```{r fig-mich-rarity}
#| fig-cap: "Every pollen taxon in the `r dplyr::n_distinct(allSamp$siteid)` Michigan sites, by the number of sites and the number of samples it occurs in. Most taxa are rare. Note that the y-axis is logarithmic."
#| fig-height: 4
ggplot(mich_taxa, aes(x = sites, y = samples)) +
  geom_point(alpha = 0.5, size = 1.6) +
  scale_y_log10(labels = label_comma()) +
  labs(x = "Number of sites the taxon occurs in",
       y = "Number of samples the taxon occurs in") +
  theme_minimal()
```

Most taxa are rare: nearly half of the taxa (`r sum(mich_taxa$sites == 1)`) occur at a single site, and only a handful (`r sum(mich_taxa$sites == max(mich_taxa$sites))`) occur at all sites. This demonstrates a core ecological concept: that common taxa are rare! 


In the stratigraphic plot demonstration we picked the top 6 species by count (purely for demonstration). @fig-mich-rarity demonstrates why. Ecologically, we would not want to throw out rare species! Such simplicity does not take into account important considerations such as functional importance, that total abundance by count may be skewed by taxa that produce a lot of pollen, and total abundance does not reflect compositional change over time.

<!-- NOTE (Claude): the shape of @fig-mich-rarity is the standard occupancy-abundance
     relationship, and the tail at sites = 1 is where identification effort and analyst
     differences live as much as ecology. Worth a sentence if you want it. -->

### How much data does each site have?

Palaeoecological datasets differ enormously in how many samples were taken and what period they cover.
For any multi-site comparison it is important to know the sampling intensity and temporal coverage.
The following code loads up the samples data from the Michigan sites, deals with the duplicate name ("Green lake"), and creates a summary object `site_cover` that checks how many samples a site has, and whether the dataset covers the whole of the Holocene period.

```{r visMichCoverage}
allSamp <- samples(mich_dl) %>%
  dplyr::filter(ecologicalgroup %in% c("UPHE", "TRSH"),
                elementtype == "pollen",
                units == "NISP") %>%      # the units filter matters, see @sec-neotoma2
  mutate(variablename = replace(variablename,
                                stringr::str_detect(variablename, "Pinus.*"),
                                "Pinus"))

dplyr::n_distinct(allSamp$siteid)

# two Michigan sites are both called "Green Lake", so the name alone is not a key
# For analyses, use siteid (not sitename), which is unique per site
# Below we do a little tidying so that the figure displays the site name
# alongside the ID for the common case where two sites have the same name!
mich_samp <- allSamp %>%
  distinct(siteid, sitename, sampleid, age, age_range_young, age_range_old) %>%
  group_by(sitename) %>%
  mutate(site_label = if (dplyr::n_distinct(siteid) > 1)
                        paste0(sitename, " (", siteid, ")") else sitename) %>%
  ungroup()

site_cover <- mich_samp %>%
  group_by(siteid, site_label) %>%
  summarise(n_samples       = n(),
            age_range_young = min(age_range_young),
            age_range_old   = max(age_range_old), .groups = "drop") %>%
  mutate(spans = age_range_young <= 0 & age_range_old >= 11700)

head(site_cover$n_samples, n = 20)
```

We can `left_join` the summary table (`site_cover`) to the original samples table so that we can use the summary information in combination with the samples data and make a new plot. We are also going to convert the `site_label` to a `factor` ordered by the number of samples, so that the plot will display the sites in order of the number of samples, rather than alphabetically.


```{r mitch-join}
mich_samp_join <- mich_samp %>%
  left_join(dplyr::select(site_cover, siteid, spans), by = "siteid") %>%
  mutate(site_label = factor(site_label,
                             levels = site_cover$site_label[order(site_cover$n_samples)]),
         ypos       = as.numeric(site_label))
head(mich_samp_join)
```


This type of data wrangling (filtering, pivoting, splitting, summarising, re-joining...) is very common in data analyses and creating visualisations. It is worth getting to grips with conceptually, to help you understand what can be done with data. Now that we have done some data wrangling, let's make a figure!


```{r fig-mich-coverage}
#| fig-cap: "Every Michigan site, ordered by how many samples it holds. Each tick is one sample. Dashed red line marks the approximate start of the Holocene at 11,700 cal years BP."
#| fig-height: 5.5

ggplot(mich_samp_join, aes(x = age, y = site_label)) +
  geom_line(aes(group = site_label), colour = "grey80", linewidth = 0.4) +
  geom_segment(aes(x = age, xend = age, y = ypos - 0.32, yend = ypos + 0.32),
                   linewidth = 0.35) +
  geom_vline(xintercept = c(11700), linetype = "dashed", colour = "red") +
  scale_x_reverse(name = "Age (cal years BP)", labels = label_comma()) +
  labs(y = NULL) +
  theme_minimal() +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        axis.text.y = element_text(size = 7))
```

@fig-mich-coverage provides a great visual summary of the Michigan data, each row mirrors the core itself! We can see from the vertical dashed red line which datasets cover the whole Holocene period (or more), and we can see the sampling intensity per site. Irwin Smith Bog was intensively sampled but does not extend over the Holocene, O'Brien Lake covers the Holocene but consists of few samples, and Bondi Section (appropriately named) holds 15 samples spanning barely a century, all of them pre-Holocene.

### Curating a comparable subset of sites

If we want to only include sites that have Holocene coverage we can use `age_range_young` and `age_range_old` to filter the metadata. We already created the `site_cover` table that has a `TRUE`/`FALSE` column that identifies which sites cover the Holocene period, so it is easy to use that to filter `siteid`s.

```{r visMichFilter}
spanning <- site_cover %>%
  dplyr::filter(age_range_young <= 0, age_range_old >= 11700)

mich_holo <- allSamp %>% dplyr::filter(siteid %in% spanning$siteid)

cat("All Michigan sites: ", dplyr::n_distinct(allSamp$siteid), "\n",
    "All Michigan samples ", dplyr::n_distinct(allSamp$sampleid), "\n",
    "Sites filtered to span the Holocene ", dplyr::n_distinct(mich_holo$siteid), "\n",
    "Total Holocene samples ", dplyr::n_distinct(mich_holo$sampleid), "\n")
```

This demonstrates a common challenge in palaeoecology: curation is expensive! Filtering by temporal extent alone removes more than half the data. If we were to go on to filter by additional criteria such as available covariates, chronological controls (i.e., a minimum number of radio carbon dates), and sampling intensity, we would lose many more sites. 

::: {.callout-warning}
# Oh no! :scream:

`age_range_young` and `age_range_old` describe the **dataset as uploaded**, not the samples left
after filtering, and the two can differ by a great deal. Restricting to upland herbs, trees and
shrubs counted as `NISP` throws away whatever else the core held, so the age range you are left
with is usually narrower than the range the metadata advertises. Check the samples you kept
rather than trusting the metadata:

```{r visMichCheck}
mich_holo %>%
  group_by(sitename) %>%
  summarise(metadata_oldest = max(age_range_old),
            oldest_kept     = round(max(age)), .groups = "drop") %>%
  mutate(difference = metadata_oldest - oldest_kept) %>%
  arrange(desc(difference))
```

Pinhook Bog advertises a record reaching past 38,000 years, while the pollen we kept stops at
about 22,000. Note also that the difference runs both ways: Wintergreen Lake's samples reach
*older* than its metadata claims.
:::


### Comparing multisite data

Because each core has its own sampling intensity and age-depth model, observations from two or more sites will not align in time. A common approach to get around this issue is to bin data at a given resolution so that samples across datasets occur within the same bin. This is quite a dangerous approach given chronological uncertainties and the effects of different bin widths. Narrow bins respect the sampling resolution but leave gaps; wide bins fill the gaps
by averaging over more time than a sample represents. Neither respects chronological uncertainty.
Two hundred years is near the fine end for this dataset. Sometimes the chronological uncertainty around a sample is larger than the bin itself! Let's explore visualising the number of samples in binned data; we won't attempt to tackle chronological uncertainties here. The following code creates 200-year bins and summarises the data by the number of samples in each bin:

```{r visMichBins}
per_site_bin <- mich_holo %>%
  dplyr::filter(age >= 0, age < 11700) %>%
  distinct(siteid, sitename, sampleid, age) %>%
  mutate(age_bin = round(age / 200) * 200) %>%
  count(age_bin, siteid, sitename, name = "n_samples")

```

Now we can make a figure that shows how many bins have how many samples aggregated, and how many bins are empty.

```{r fig-mich-bins}
#| fig-cap: "Samples per 200 year bin for each site that spans the window. White cells are bins where that site has no sample at all, and they are a third of the grid."
#| fig-height: 4
per_site_bin %>%
  mutate(sitename = factor(sitename,
           levels = names(sort(tapply(n_samples, sitename, sum))))) %>%
  ggplot(aes(x = age_bin, y = sitename, fill = n_samples)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Samples in bin", option = "mako", direction = -1) +
  scale_x_reverse(name = "Age (cal years BP)", labels = label_comma()) +
  labs(y = NULL) +
  theme_minimal() +
  theme(panel.grid = element_blank(), legend.position = "top",
        axis.text.y = element_text(size = 7))
```



### Where are taxa present?

Bearing in mind that we have glossed over the quality of each site's chronology, it is still worth looking at another plot to demonstrate how data from Neotoma can be explored. The following code once again filters for the most abundant six taxa and records that cover the Holocene, follows the same binning procedure shown above, and creates a table of how many sites record each taxon in each bin.

```{r visPresence}
#| output: false
taxa_by_age <- mich_holo %>%
  dplyr::filter(variablename %in% focal, age >= 0, age < 11700) %>%
  group_by(variablename, age_bin = round(age / 200) * 200) %>%
  summarise(n_sites_with_taxon = dplyr::n_distinct(siteid), .groups = "drop")

sites_by_age <- mich_holo %>%
  dplyr::filter(age >= 0, age < 11700) %>%
  group_by(age_bin = round(age / 200) * 200) %>%
  summarise(n_sites_total = dplyr::n_distinct(siteid), .groups = "drop")

presence <- taxa_by_age %>%
  inner_join(sites_by_age, by = "age_bin") %>%
  mutate(proportion   = n_sites_with_taxon / n_sites_total,
         variablename = factor(variablename, levels = focal))
```

Before plotting it, look at a summary table:

```{r visPresenceCheck}
presence %>%
  group_by(variablename) %>%
  summarise(bins = n(),
            min_prop = min(proportion),
            max_prop = max(proportion),
            bins_at_100pct = sum(proportion == 1),
            .groups = "drop")
```

Now let's make a plot:

```{r fig-mich-presence}
#| fig-cap: "Proportion of the 13 curated Michigan sites recording each taxon through time, in 200 year bins, with a binomial GAM smooth weighted by the number of sites contributing to each bin. Four of the six taxa sit at or near 100% throughout, which is the problem described below."
#| fig-height: 4.5
#| warning: false

ggplot(presence, aes(x = age_bin, y = proportion, weight = n_sites_total)) +
  geom_point(size = 1) +
  geom_smooth(method = "gam", formula = y ~ s(x),
              method.args = list(family = "binomial")) +
  facet_wrap(~variablename,
             labeller = as_labeller(taxon_names)) +
  scale_x_reverse(name = "Age (cal years BP)", labels = label_comma()) +
  scale_y_continuous(name = "Proportion of sites recording the taxon",
                     labels = label_percent()) +
  theme_minimal() +
  theme(strip.text = element_marquee())
```

Despite making several assumptions along the way about chronological controls and bin widths (it is advisable to experiment with several different bin widths), some patterns do emerge!
*Quercus* and *Ulmus* are present at every site in every one of the 56 age bins, and *Pinus* in 48 of
them. The smoother is purely a visual guide, not intended for statistical analyses. Presence is only informative for a taxon that is sometimes absent, which here is *Picea*, which ranges from 9% to 100% of sites, and *Fraxinus*, which never exceeds half of them.


::: {.callout-note}
# Always remember ecological knowledge! :mage:

_Always_ remember the ecology of a region. It is one thing to play with the numbers from a database, but we could be missing important ecological dynamics. The above example visualises data across Michigan as one, so this approach could be missing those important underlying dynamics. Michigan is split by a climatic tension zone. How would the results change if that knowledge, which is not contained in the downloads, were included?

:::


::: {#exr-youchoose}
We have worked through several examples of how data from Neotoma can be manipulated and reshaped to produce data visualisations. This exercise is an open-ended one: either take one of the visualisations above, improve it, and discuss the shortcomings and what issues you addressed, _or_ come up with a visualisation of your own and discuss what you did and why.

There are still many possible visualisations. If you need some inspiration, consider:

- Visualising chronological uncertainty
- Spatial change in taxa (e.g., longitude against abundance)
- Species richness against counting effort
- Some sites include the name of the analyst under `sampleanalyst`. Which sites have multiple analysts, and does anything change in the data depending on who the analyst was?

:::


## Conclusion

This chapter is intended to: (i) give you insight into how many ways there are to manipulate and visualise data for different purposes; (ii) highlight pitfalls and potential errors such as ignoring chronological uncertainty or sampling intensity, and also missing ecological knowledge that the database does not contain directly. Some datasets may have associated publications that you can read to understand the methods and regional history. These points are far more important than the code itself which can now be written in minutes by AI. If you are using AI, make sure that every line of code is verified and you know exactly what the output is and how it was created. I take the approach that there are no AI mistakes, only my own misdirection or lack of verification.

The exercises are worth returning to if you skipped them:

- @exr-freeaxis, what the `scales` argument does to a stratigraphic plot
- @exr-downsample, what falls away as temporal resolution drops
- @exr-youchoose, improve on one of the examples or come up with a new visualisation
