# Visualising pollen {#sec-visualisation}

## Background

<!-- NOTE (Quinn's, kept): include the issue of visualising relative proportions. -->
<!-- NOTE (Quinn's, kept): I use segment in ggplot to indicate where observations lie. Many
     publications do not do this but it is important to see where the sampling intervals lie,
     otherwise you have an automatic linear interpolation! This is now the point that
     @fig-strat-devils makes, and the first plot below deliberately omits the segments so the
     difference is visible. -->
<!-- NOTE (Claude): this section still needs the real Background prose. The shape it wants is
     probably: what a pollen diagram is and why it is drawn the way it is (time vertical, oldest
     at the bottom, one panel per taxon), then the two honesty problems the chapter deals with,
     which are (a) proportions hide the count they came from and (b) an area or line between two
     samples is an interpolation the data do not support. -->

A pollen diagram plots each taxon in its own panel against depth or age, with the oldest
material at the bottom, mirroring the sediment core it came from. This chapter builds one twice,
once by hand with `ggplot2` so that every choice is visible, and once with `rioja`, which is the
package most palynologists reach for.


## Getting started

::: {.callout-note}
# Packages required for this section

```{r visPackages}
#| output: false
if (!require("pacman")) install.packages("pacman", repos = "https://cran.r-project.org")
pacman::p_load(dplyr, tidyr, tibble, ggplot2, scales, stringr, mgcv, rioja, neotoma2, marquee)
```
:::

::: {.callout-tip}
# Respawn code

This is the same Devil's Lake download used in @sec-age-depth and @sec-dissimilarity, filtered
to `datasettype == "pollen"`, harmonised the same way, and converted to proportions.

```{r visDevilsDownload}
#| output: false
if (!file.exists("./data/devils_dl.rds")) {
  devils_dl <- get_sites(siteid = 666) %>%
    get_datasets(all_data = TRUE) %>%
    neotoma2::filter(datasettype == "pollen") %>%
    get_downloads()
  saveRDS(devils_dl, "./data/devils_dl.rds")
} else {
  devils_dl <- readRDS("./data/devils_dl.rds")
}

```
:::


Most stratigraphic plots you will see have taxa converted to relative abundances. The following code converts observations from counts (NISP) to proportions. In this section we will not go into detail on the `samples()` function or the conversion to proportions as this is covered in @sec-neotoma2 and @sec-dissimilarity.

```{r ConvertToProp}
devils_samples <- samples(devils_dl) %>%
  dplyr::filter(ecologicalgroup %in% c("UPHE", "TRSH"),
                elementtype == "pollen",
                units == "NISP")

devils_harm_prop <- devils_samples %>%
  mutate(variablename = replace(variablename,
                                stringr::str_detect(variablename, "Pinus.*"),
                                "Pinus")) %>%
  group_by(siteid, sitename, sampleid, variablename, units,
           age, agetype, depth, datasetid, long, lat) %>%
  summarise(value = sum(value), .groups = "drop") %>%
  group_by(age) %>%
  mutate(pollencount = sum(value, na.rm = TRUE),
         prop        = value / pollencount) %>%
  ungroup() %>%
  arrange(desc(age))
```



## Stratigraphic diagrams

Here, we will take the Devil's Lake data and make some statigaphic diagrams comparable to those seen in publications.

Six taxa are enough to show the method. They are chosen because they span the record between
them rather than because they are the six most abundant.

```{r visFocalTaxa}
focal <- c("Pinus", "Picea", "Quercus", "Betula", "Tsuga", "Ulmus")

devils_focal <- devils_harm_prop %>%
  dplyr::filter(variablename %in% focal)

```

### Building one with `ggplot2`

Start with the simplest thing that works: proportion against age, one panel per taxon.
`coord_flip()` puts age on the vertical axis and `scale_x_reverse()` makes it run downwards, so
the oldest sample sits at the bottom.

```{r fig-strat-first}
#| fig-cap: "A first pollen diagram for Devil's Lake. Readable, and misleading in two ways that the next version fixes."
#| fig-height: 5
ggplot(devils_focal, aes(x = age, y = prop)) +
  geom_area(fill = "grey85") +
  scale_x_reverse(name = "Age (cal years BP)", labels = label_comma()) +
  scale_y_continuous(name = "Proportion of the pollen sum", labels = label_percent()) +
  coord_flip() +
  facet_wrap(~variablename, nrow = 1) +
  theme_minimal()
```

Two things are wrong with @fig-strat-first, and neither is cosmetic.

The filled area implies a continuous record. It is not: the core was sampled at intervals, and
everything between two samples is drawn by the plotting function, not measured. Adding a segment
at each sampled age shows where the data actually are.

The panels are also in alphabetical order, which carries no information. Ordering them by the
weighted average of the ages at which each taxon occurs puts the taxa in the order they
dominated the record, which is what `rioja::strat.plot()` does with `wa.order`.

```{r visWaOrder}
wa_order <- devils_focal %>%
  group_by(variablename) %>%
  summarise(wa_age = sum(age * prop) / sum(prop), .groups = "drop") %>%
  arrange(desc(wa_age))

wa_order
```

```{r visFactorOrder}
devils_focal <- devils_focal %>%
  mutate(variablename = factor(variablename, levels = wa_order$variablename))
```

#### Formatting taxon names

Genus and species names are italic; families, orders and qualifiers such as `undiff.` are not.
Doing that inside a facet strip means handing `ggplot2` a plotmath expression rather than a
plain string, which `label_parsed` then renders.

<!-- NOTE (Claude): you asked about the marquee package. Neither marquee nor ggtext is installed
     here, and both would be a new dependency in install.R. plotmath needs nothing extra and is
     what is used below. the element_markdown() function in ggtext is the usual alternative and reads more
     easily (`*Pinus* undiff.`), so if you would rather have that, say and I will swap it and add
     ggtext to install.R. -->

```{r visTaxonLabel}
# Turn a Neotoma taxon name into a plotmath expression.
# Genus and species italic; family and above, and every qualifier, roman.
taxon_label <- function(x) {
  vapply(x, function(nm) {
    qual <- ""
    if (grepl(" undiff\\.$", nm)) {
      qual <- "undiff."; nm <- sub(" undiff\\.$", "", nm)
    } else if (grepl("-type$", nm)) {
      qual <- "-type";   nm <- sub("-type$", "", nm)
    } else if (grepl(" sensu (lato|stricto)$", nm)) {
      qual <- sub("^.* (sensu (?:lato|stricto))$", "\\1", nm)
      nm   <- sub(" sensu (lato|stricto)$", "", nm)
    }

    body <- if (grepl(" subg\\. ", nm)) {                  # Pinus subg. Strobus
      p <- strsplit(nm, " subg\\. ")[[1]]
      sprintf('italic("%s")~plain("subg.")~italic("%s")', p[1], p[2])
    } else if (grepl("aceae$|ales$|phyta$|opsida$|ae$", nm)) {   # family and above
      sprintf('plain("%s")', nm)
    } else {                                               # genus or species
      sprintf('italic("%s")', nm)
    }

    if (qual == "")           body
    else if (qual == "-type") paste0(body, '*plain("-type")')
    else                      paste0(body, '~plain("', qual, '")')
  }, character(1), USE.NAMES = FALSE)
}
```

Worth checking against the awkward cases before trusting it, because a malformed expression is a
render error rather than a wrong-looking label:

```{r visTaxonLabelCheck}
check <- c("Pinus", "Asteraceae", "Pinus subg. Strobus",
           "Ambrosia-type", "Asteraceae undiff.", "Caprifoliaceae sensu lato")

tibble::tibble(name = check, plotmath = taxon_label(check))
```

`as_labeller()` takes a named vector of those expressions, and `default = label_parsed` tells
`ggplot2` to parse rather than print them.

```{r visLabelMap}
strip_labels <- setNames(taxon_label(levels(devils_focal$variablename)),
                         levels(devils_focal$variablename))
```

Now the same plot with the segments, the ordering and the names.

```{r fig-strat-devils}
#| fig-cap: "Devil's Lake pollen diagram. Each horizontal segment is a sampled level, so the width of the shaded area between segments is interpolation rather than data. Panels are ordered by the weighted average age at which each taxon occurs, oldest on the left."
#| fig-height: 5
ggplot(devils_focal, aes(x = age, y = prop)) +
  geom_area(fill = "grey85") +
  geom_segment(aes(x = age, xend = age, y = 0, yend = prop),
               colour = "grey30", linewidth = 0.3) +
  scale_x_reverse(name = "Age (cal years BP)", labels = label_comma()) +
  scale_y_continuous(name = "Proportion of the pollen sum",
                     labels = label_percent(), n.breaks = 3) +
  coord_flip() +
  facet_wrap(~variablename, nrow = 1, scales = "free_x",
             labeller = as_labeller(strip_labels, default = label_parsed)) +
  theme_minimal() +
  theme(panel.spacing.x   = unit(0.9, "lines"),
        panel.grid.minor  = element_blank(),
        axis.text.x       = element_text(size = 7))
```

<!-- NOTE (Claude): `scales = "free_x"` gives every taxon its own proportion axis, which is the
     only way Tsuga at about 1% and Quercus at about 60% are both visible. The cost is that panel
     widths are no longer comparable by eye, which is exactly what a reader is tempted to do.
     rioja solves this differently, with a common scale plus an exaggeration factor. Worth a
     sentence deciding which convention the book teaches. -->

### The same diagram with `rioja`

`rioja` is built for this and does in one call what the section above does in several, including
the CONISS constrained cluster analysis that splits the record into zones. It wants a wide
matrix, samples in rows and taxa in columns.

```{r visWideTable}
#| output: false
devils_wide <- devils_harm_prop %>%
  dplyr::select(age, variablename, prop) %>%
  dplyr::filter(variablename %in% focal) %>%
  tidyr::pivot_wider(id_cols = age,
                     names_from  = variablename,
                     values_from = prop,
                     values_fill = 0)

dim(devils_wide)
```

```{r fig-strat-rioja}
#| fig-cap: "The same six taxa drawn by `rioja::strat.plot()`, with a CONISS dendrogram on the right and four zones marked in red. Taxa are ordered by weighted average, as in the previous figure."
#| fig-height: 6
#| warning: false
devils_clust <- rioja::chclust(dist(sqrt(devils_wide[, -1])), method = "coniss")

strat <- rioja::strat.plot(as.data.frame(devils_wide[, -1] * 100),
                           yvar   = devils_wide$age,
                           title  = devils_samples$sitename[1],
                           ylabel = "Age (cal years BP)",
                           y.rev  = TRUE,
                           clust  = devils_clust,
                           wa.order = "topleft",
                           scale.percent = TRUE)

rioja::addClustZone(strat, devils_clust, 4, col = "red")
```

<!-- NOTE (Claude): the object is called `strat` rather than `plot`, which is what the old draft
     used. Assigning to `plot` masks base R's plot() for the rest of the session. -->

<!-- NOTE (Claude): CONISS and the distance measure behind it are @sec-dissimilarity's subject,
     and this chapter currently comes first. Either a forward reference here, or move this
     chapter after dissimilarity. Flagging rather than deciding. -->

## Taxa across space and time

One site gives a history. Many sites give a geography as well. The Michigan download from @sec-neotoma2 holds many sites, and the same harmonisation
applies to all of them.

```{r visMichSamples}
mich_dl <- readRDS("./data/mich_dl.rds")

allSamp <- samples(mich_dl) %>%
  dplyr::filter(ecologicalgroup %in% c("UPHE", "TRSH"),
                elementtype == "pollen",
                units == "NISP") %>%      # the units filter matters, see @sec-neotoma2
  mutate(variablename = replace(variablename,
                                stringr::str_detect(variablename, "Pinus.*"),
                                "Pinus"))

dplyr::n_distinct(allSamp$siteid)
```

### Where a taxon is present

The obvious first question is what fraction of sites record a taxon at a given time. Ages are
binned to the nearest 500 years, which is what `round(age * 2, -3) / 2` does.

```{r visPresence}
#| output: false
taxa_by_age <- allSamp %>%
  dplyr::filter(variablename %in% focal, age < 11000) %>%
  group_by(variablename, age_bin = round(age * 2, -3) / 2) %>%
  summarise(n_sites_with_taxon = dplyr::n_distinct(siteid), .groups = "drop")

sites_by_age <- allSamp %>%
  dplyr::filter(age < 11000) %>%
  group_by(age_bin = round(age * 2, -3) / 2) %>%
  summarise(n_sites_total = dplyr::n_distinct(siteid), .groups = "drop")

presence <- taxa_by_age %>%
  inner_join(sites_by_age, by = "age_bin") %>%
  mutate(proportion   = n_sites_with_taxon / n_sites_total,
         variablename = factor(variablename, levels = wa_order$variablename))
```

Before plotting it, look at what the numbers do:

```{r visPresenceCheck}
presence %>%
  group_by(variablename) %>%
  summarise(bins = n(),
            min_prop = min(proportion),
            max_prop = max(proportion),
            bins_at_100pct = sum(proportion == 1),
            .groups = "drop")
```

```{r fig-mich-presence}
#| fig-cap: "Proportion of Michigan sites recording each taxon through time, with a binomial GAM smooth weighted by the number of sites in each bin. Four of the six taxa are at or near 100% throughout, which is the problem described below."
#| fig-height: 4.5
#| warning: false
# warning: false suppresses two "iteration limit reached" messages from mgcv. They are a
# real symptom, not noise: a taxon present at 100% of sites in every bin is completely
# separated, so the fit cannot converge. The table above shows the same thing in numbers.
ggplot(presence, aes(x = age_bin, y = proportion, weight = n_sites_total)) +
  geom_point(size = 1) +
  geom_smooth(method = "gam", formula = y ~ s(x),
              method.args = list(family = "binomial")) +
  facet_wrap(~variablename,
             labeller = as_labeller(strip_labels, default = label_parsed)) +
  scale_x_reverse(name = "Age (cal years BP)", labels = label_comma()) +
  scale_y_continuous(name = "Proportion of sites recording the taxon",
                     labels = label_percent()) +
  theme_bw()
```

*Quercus* is present at every site in every one of the 23 bins, and *Pinus* and *Ulmus* in 22 of
them. For those taxa the measure cannot go up or down, so the smooth is flat and tells you
nothing. Presence is only informative for a taxon that is sometimes absent, which here is
*Picea* and *Tsuga*.

<!-- NOTE (Claude): two things worth saying in prose here, both of which the code above already
     demonstrates. First, `weight = n_sites_total` is not optional. Without it the binomial GAM
     is fitted to bare proportions and mgcv warns "non-integer #successes" once per panel; the
     old draft hid that with `warning: false`. Second, a taxon at 100% in every bin is complete
     separation, which is why the fit struggles. This is the honest lead-in to the abundance
     version below, which was your note in the old draft. -->

### How abundant it is

Mean proportional abundance moves even when presence does not, so it is the more useful
measure for common taxa. Proportions have to be calculated within each sample first, so that
sites counted to different totals contribute equally.

```{r visAbundance}
#| output: false
mich_abund <- allSamp %>%
  group_by(siteid, sampleid, age) %>%
  mutate(prop = value / sum(value, na.rm = TRUE)) %>%   # within-sample proportions first
  ungroup() %>%
  dplyr::filter(variablename %in% focal, age < 11000) %>%
  group_by(variablename, age_bin = round(age * 2, -3) / 2) %>%
  summarise(mean_prop = mean(prop),
            n_sites   = dplyr::n_distinct(siteid), .groups = "drop") %>%
  mutate(variablename = factor(variablename, levels = wa_order$variablename))
```

```{r fig-mich-abundance}
#| fig-cap: "Mean proportion of the pollen sum for each taxon across Michigan sites, binned to 500 years. Point size is the number of sites contributing to each bin. Unlike @fig-mich-presence, every taxon varies."
#| fig-height: 4.5
ggplot(mich_abund, aes(x = age_bin, y = mean_prop)) +
  geom_line(colour = "grey40") +
  geom_point(aes(size = n_sites), alpha = 0.6) +
  scale_size_continuous(name = "Sites", range = c(0.4, 2.4)) +
  facet_wrap(~variablename,
             labeller = as_labeller(strip_labels, default = label_parsed)) +
  scale_x_reverse(name = "Age (cal years BP)", labels = label_comma()) +
  scale_y_continuous(name = "Mean proportion of the pollen sum",
                     labels = label_percent(accuracy = 1)) +
  theme_bw()
```

<!-- NOTE (Claude): @fig-mich-abundance is where the ecology is, and the prose could point at the
     Picea decline, the mid-Holocene Tsuga rise and the Quercus maximum. The Tsuga trough at
     around 4,500 years is the well-known hemlock decline and is worth a citation. -->

<!-- NOTE (Claude): a further step, if you want one, is that binning to 500 years and taking an
     unweighted mean treats a bin holding two sites the same as one holding twenty. Point size
     shows the imbalance but does not correct it. That is a natural lead-in to @sec-gams. -->

## Exercises

::: {#exr-taxon-names}
`taxon_label()` handles the qualifiers that appear at Devil's Lake. Run it over the full taxon
list from a different site and find a name it formats wrongly.

1. Get the taxon list with `unique(samples(...)$variablename)` for a site of your choice.
2. Pass the whole list through `taxon_label()`, and check every result parses with
   `parse(text = ...)` inside `tryCatch()`.
3. Any name that parses can still be wrong. Find one that is formatted incorrectly rather than
   one that errors, say what the rule should be, and add it to the function.
:::

::: {#exr-stratplot}
Draw a stratigraphic diagram for a site other than Devil's Lake, with taxa of your choosing,
using either `ggplot2` or `rioja`.

1. Pick a site using the search methods in @sec-neotoma2 and download it with the gated pattern
   in the respawn code above.
2. Choose between five and eight taxa. Say why you chose those, given that the diagram is meant
   to show how the site changed.
3. Order the panels by weighted average age. Does the order match what you expected from the
   ecology?
:::

::: {#exr-taxon-region}
Repeat @fig-mich-abundance for a different state or region.

1. Download the sites for your region, following @sec-neotoma2.
2. Produce both the presence and the abundance version.
3. Report how many of your taxa are present at 100% of sites in every bin. If the answer is most
   of them, what does that tell you about the taxa you picked rather than about the region?
:::

## Conclusion

<!-- NOTE (Claude): placeholder, yours to write. The three things the chapter actually
     establishes are: a pollen diagram is a set of per-taxon panels against age with the oldest
     at the bottom; anything drawn between two sampled levels is interpolation and should be
     shown as such; and a measure that cannot vary, such as presence for a taxon found
     everywhere, will produce a flat line that looks like a result. -->

The exercises are worth returning to if you skipped them:

- @exr-taxon-names, formatting taxon names correctly
- @exr-stratplot, a stratigraphic diagram of your own
- @exr-taxon-region, presence against abundance in another region
